---
title: "DTSC-550: Getting Started with R"
output: pdf
---

# VIDEO: Before we get started in R & RStudio

Before you can get started, do these:

1.  Install R on your computer

2.  Install RStudio on your computer

This [link](https://posit.co/download/rstudio-desktop/) provides an option to download both R and R Studio.

3.  Download this `.rmd` file from Brightspace (use "Files" to open it)

4.  Access the "Learning Statistics with R" [textbook](https://learningstatisticswithr.com/lsr-0.6.pdf) for additional help.

Much of the following comes from [the textbook](https://learningstatisticswithr.com/book/introR.html)

5.  Know the difference between R and R Studio

6.  Things to know about R Studio

-   Source pane

-   Console

-   Environment pane

-   Plot/output pane

-   Insert chunk; run chunk

```{r}

# to run on a Mac: command + return

```

```{r}
# On a Mac: option + command + i gives a new code chunk

# On a PC: control + alt + i 
```

-   Source vs. Visual

-   `#` inside block is a comment; `#` outside a block creates the outline

-   Outline

7.  Under "Help", see "Cheatsheets"

------------------------------------------------------------------------

# VIDEO: Using R as a Calculator

Typing commands at the R Console vs. in a code block

```{r}
1+10
```

# You don't need the initial `>` sign

```{r}
# comment: this will yield an error message!
 10 + 20
```


```{r}
1+1
```

How did I make this "code chunk"?

# Be very careful to avoid typos

## Typo: You meant `+` but typed `=`

```{r}
# This will yield an error!
10 = 20

```

## Typo: You meant `20 - 10`

```{r}
10 - 20
```

# R is flexible with spacing, but use common practice!

Especially on CodeGrade assignments and the "structure tests"

## Example

```{r}
# Don't do this
10        + 20

# Don't do this either
10+20

# Recommended
10 + 20
```

## Another example with spaces

```{r}
citation()
```

```{r}
### This works, but not recommended
citation (  )
```

```{r}
### This doesn't work!
### Notice the red X to the left
cit ation()
```

# R can sometimes tell that you're finished (but not often)

```{r}
10 +
```

```{r}
10 + (3 * 4

### Some calculators are ok with this...not R!
```

# Math Operations

```{r}
### note the recommended spacing

10 + 2 
10 - 2
10 * 2
10 / 2
10 ^ 2

# What's with the [1]?
```

# Order of Operations: PEMDAS (or BEMDAS)

```{r}
(1 + 2) * 3

#[1 + 2] * 3

1 + 2 * 3
```

------------------------------------------------------------------------

# VIDEO: The Assignment Operator in R

## Storing a number as a variable

`<-` is the assignment operator in R.

You could use `=`, but it's not common practice.

```{r}
sales <- 350

#sales = 350

### Note environment pane!

sales
print(sales)
```

Just because you store a number into a variable doesn't mean you can see it immediately. You still need to `print()` or display it.

## Don't assign values like this!

```{r}
### These work but aren't common practice
sales<-350

350 -> sales

sales = 350

### This doesn't work
sales < - 350
```

## Doing calculations with variables

```{r}
sales <- 350
royalty <- 7

350*7

sales * royalty
```

```{r}
revenue <- sales * royalty
revenue

### Note the spacing
```

## Using `()` to print

-   This isn't common practice

-   Avoid this on assignments!

```{r}
(sales * royalty)
```

## Overwrite old value with new value

Be careful when you do this, especially when you're working with large files...you don't want to accidentally overwrite your work.

Note DTSC-580 (Data Manipulation) course

```{r}
revenue <- revenue + 550
```

------------------------------------------------------------------------

# VIDEO: Rules for naming variables in R

-   "Run all chunks above" option (in 2 places)

## 1. You can use uppercase A-Z, lowercase a-z, numbers 0-9 and punctuation `.` and `_` in the variable name.

```{r}
### We don't recommend this!
SaL.e_s <- 100
```

```{r}
Sales? <- 100
```

## 2. Variable names cannot include spaces...consider using a `.` instead

```{r}
sa les <- 100
```

```{r}
my.sales <- 100
```

## 3. Variable names must start with a letter or a period, not an underscore or a number. Convention: don't start with a period, even though it's allowed.

```{r}
sales <- 100

Sales <- 101

.sales <- 50 ### not recommended

```

```{r}
# error!
_sales <- 100
```

```{r}
# error!
1718sales <- 2000
```

```{r}
sales1718 <- 2000
```

### Errors

```{r}
# Error!
_sales <- 50
1sales <- 50
```

## 4. Variable names cannot be one of the reserved keywords, such as `if`, `else`, `repeat`, `while`, `function`, `for`, `in`, `next`, `break`, `TRUE`, `FALSE`, `NULL`, `Inf`, `NaN`, `NA`, etc.

```{r}
# error!
break <- 999
```

You don't need to memorize these! You'll know it when you break it.

## Conventions for Naming Variables

### 1. Use informative variable names

Good: sales & revenue

Bad: var1, var2, df1, df2

### 2. Use short names

```{r}
sales.for.this.book.that.you.are.reading <- 1000000

#sales.for.this.book.that.you.are.reading
```

-   Typing is a pain
-   Reading is a pain
-   Tension between informative & concise

### 3. Be consistent with your variable naming style

-   `my.new.salary` (aka dot case or period.separated)

-   `my_new_salary` (aka snake case)

-   `myNewSalary` (aka camel case)

-   `my-new-salary` (aka kebab case...not allowed in R)

-   `MyNewSalary` (aka Pascal case)

-   Check with your organization if they have a style preference

-   Different coding languages have different style recommendations/requirements

### 4. For CodeGrade assignments, use the names we ask you to use

-   We typically have you use Q1, Q2, Q3, etc. for your variable (object) names

------------------------------------------------------------------------

# VIDEO: More Math Operations in R

## Math Operations

```{r}
### Square Root
225 ^ 0.5

sqrt(225)

### Absolute Value
abs(21)

abs(-21)

### Multiple Functions
sqrt(1 + abs(-8))
```

### R works with Real Numbers only...

```{r}
sqrt(-9) # NaN = not a (real) number
```

### ...but Complex numbers are possible

But we won't be using these in this course (or this program)!

```{r}
sqrt(-9+0i)
```

## Function arguments, their names and their defaults

```{r}
round(3.1415)

round(3.14165, 2)

round(x = 3.1415, digits = 2)

### Don't do this, even though it works
### We only use <- to create new variables
round(x <- 3.1415, digits <- 3)

### Again, don't use <- like this!

```

```{r}
round(3.14165, 2)

### The order of the arguments matter!
round(2, 3.14165)
```

------------------------------------------------------------------------

# VIDEO: Autocomplete, Tab Completion, History

## Autocomplete & Tab completion

These work here in a notebook *and* in the console

```{r}
### With a built-in function, like round



### With the functions arguments



### With the name of a saved variable, like sales from earlier

```

## Browsing your command history

1.  In console, hit up arrow
2.  In R Studio, find "history" in the Environment pane (upper right)

------------------------------------------------------------------------

# VIDEO: Vectors and Operations

## Storing many numbers as a vector

Suppose the sales of a textbook were 0 books in January, 100 in February, 200 in March, 50 in April, and none the rest of the year.

Use the "combine" function (or collection or concatenate):

```{r}
sales.by.month <- c(0, 100, 200, 50, 0, 0, 0, 0, 0, 0, 0, 0)

sales.by.month

### Adjust window to discuss indices 
```

## Getting information out of vectors

```{r}
sales.by.month[2]
```

Note that `[1] 100` is what prints, not `[2] 100`.

```{r}
feb.sales <- sales.by.month[2]
```

## Using `:` for a range of indices

```{r}
### Sales the last half of the year: July through Dec.
sales.by.month[7:12]
```

## Altering the elements of a vector

```{r}
sales.by.month[5] <- 25

sales.by.month
```

Python begins with 0 as its first index; R uses 1.

Note the codeblock above overwrites the `sales.by.month` variable.

## You can't index what doesn't exist

```{r}
sales.by.month[13]
```

## Other useful functions with vectors

### length()

```{r}
length(sales.by.month)
```

### Vectorized operations

```{r}
sales.by.month * 7
```

```{r}
days.per.month <- c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31)

profit <- sales.by.month * 7

profit / days.per.month
```

------------------------------------------------------------------------

# VIDEO: Storing Other Types of Data in R

## Storing text data

```{r}
greeting <- 'hello'
greeting

greeting2 <- "hello"
greeting2
```

```{r}
months <- c("January", "Feb", "March", "April", "May", "June", "July", "Aug", "Sept", "Oct", "Nov", "Dec")

months[12]

months[c(2, 3, 4, 5)]

months[2:5]
```

### length() vs. nchar()

```{r}
length(greeting)

nchar(greeting)

length(months)

nchar(months)
```

## Storing "true or false" data

```{r}
2 + 2
```

```{r}
2 + 2 == 4
```

```{r}
2 + 2 == 5
```

```{r}
2 + 2 = 4
```

## Logical operations

```{r}
3*3 + 4*4 == 5*5
```

```{r}
sqrt(25) == 5
```

```{r}
99 < 100
```

```{r}
100 < 100
```

```{r}
100 <= 100
```

```{r}
25 != 30
```

## Vectors of Logicals

```{r}
x <- c(TRUE, TRUE, FALSE)

length(x)

sum(x)

x == 1

x == 0
```

------------------------------------------------------------------------

# VIDEO: Indexing Vectors

## How do we access our values stored in a vector?

```{r}
sales.by.month[c(2, 3, 4)]

sales.by.month[2:4]

```

```{r}
sales.by.month[4:2]

sales.by.month[c(4, 3, 2)]
```

```{r}
sales.by.month

sales.by.month[c(2, 7, 12)]
```

## It's easy to forget the `c()`!

```{r}
sales.by.month[2, 7, 12]
```

## Logical indexing

Which months had positive sales (i.e., more than \$0)?

```{r}
sales.by.month
```

Note: It may seem easy to do this by eye, but with millions of rows, we need something more efficient (and less error-prone) than our counting

```{r}
sales.by.month > 0
```

## The following is super important when indexing with logical operators!

### Do we want the sales themselves?

```{r}
test <-sales.by.month[sales.by.month > 0]
```

### Or, do we want the months?

```{r}
months[sales.by.month > 0]
```

For the months that had positive sales, how much were those sales?

```{r}
sales.by.month[sales.by.month > 0]
```

------------------------------------------------------------------------

# VIDEO: Global Options and Set Working Directory

## Get working directory

```{r}
getwd()
```

## See the files in your working directory

```{r}
list.files()
```

## Set working directory

```{r}
setwd("/Users/home/Desktop")

### Syntax differs between PC & MAC
```

## Global option to set working directory

1.  Tools
2.  Global Options
3.  Under Basic tab, select "Browse" to set your default working directory

See textbook discussion [here](https://learningstatisticswithr.com/book/mechanics.html#navigation)

## How to quit R

```{r}
q()
```

Save: Do you want to save all the variables in the environment?

------------------------------------------------------------------------

# VIDEO: Using Comments

`#` tells R to ignore everything else you've written on this line, after the `#`

It's tempting to skip comments, especially when your code isn't that complicated.

But, as your code gets more complicated, using comments is essential. Your teammates and your future self with thank you.

```{r}
seeker <- 3.1415           # create the first variable 5*5
lover <- 2.7183            # create the second variable
keeper <- seeker * lover   # now multiply them to create a third one
print(keeper)              # print out the value of 'keeper'
```

------------------------------------------------------------------------

# VIDEO: Packages

A *package* is basically just a big collection of functions, data sets and other R objects that are all grouped together under a common name.

Some packages are already installed when you put R on your computer, but the vast majority of R packages are out there on the internet, waiting for you to download, install and use them.

These are called "libraries" in Python, such as NumPy, Pandas, Seaborn.

------------------------------------------------------------------------

There's a critical distinction that you need to understand: the difference between having a package installed on your computer, and having a package loaded in R.

1.  A package must be *installed* before it can be loaded.

2.  A package must be *loaded* before it can be used.

Citation [here](https://learningstatisticswithr.com/book/mechanics.html#packageinstall)

Light Bulb Analogy:

-   Installing a package is like installing a light bulb

-   Load a package is like turning on a light bulb

------------------------------------------------------------------------

Let's check out the packages panel to understand the difference between packages that are *installed* and ones that are *loaded*.

## Either the package isn't installed or it isn't loaded...

```{r}
dplyr
```

## 1. How to install a package

You only do this once.

```{r}
install.packages("dplyr")
```

## 2. How to load the package

You do this each time you use want to use the package, typically at the beginning of your notebook.

```{r}
library(dplyr)
```

## Let's practice installing

```{r}
install.packages("psych")
```

## ...and let's practice loading

```{r}
library(psych)

### Note the lack of quotes around psych
```

------------------------------------------------------------------------

## Important Note about R Tools!

RTools is a package to help build and develop packages, which we won't do in this course.

If you are running Windows OS, you may see a message that says: *WARNING: Rtools is required to build R packages, but is not currently installed.* You can safely ignore this warning.

------------------------------------------------------------------------

## Update packages

There's an `update.packages()` function but it's easier to use the packages panel and select "Update".

The [LSR textbook](https://learningstatisticswithr.com/book/mechanics.html#what-packages-does-this-book-use) uses these packages:

-   lsr

-   psych

-   car

```{r}
#install.packages(c('lsr', 'psych', 'car'))

library(lsr)
library(psych)
library(car)
```

------------------------------------------------------------------------

# VIDEO: Cleaning up your environment

How do I clear my environment?

## Use rm() for remove

```{r}
### Remove one object/variable
rm(greeting)
```

```{r}
### Remove all objects
rm(list = ls())
```

## Use the broom buttons

-   in the environment pane

-   in the console

------------------------------------------------------------------------

# VIDEO: Intro to Data frames

For more, see [textbook](https://learningstatisticswithr.com/book/mechanics.html#dataframes)

## Example

Suppose an experiment was conducted with three treatment groups and yielded the following data:

```{r}
# Treatment group
group <- c(1, 1, 1, 2, 2, 2, 3, 3, 3)

# Gender of participants
gender <- c('male', 'male', 'male', 'male', 'male', 'female', 'female', 'female', 'female')

# Ages of participants
age <- c(17, 19, 21, 37, 18, 19, 47, 18, 19)

# Scores on a test
score <- c(12, 10, 11, 15, 16, 14, 25, 21, 29)
```

All four of these variables correspond to the same data set, and all four of them are organised in exactly the same way...but R doesn't know this!

In other words, when we store everything in separate variables like this, R doesn’t know anything about the relationships between things. It doesn’t even really know that these variables actually refer to a proper data set.

The data frame fixes this: if we store our variables inside a data frame, we’re telling R to treat these variables as a single, fairly coherent data set.

## Create data frame

```{r}
expt <- data.frame(age, gender, group, score) 
expt 
```

------------------------------------------------------------------------

# VIDEO: Pulling out contents from a data frame

We use a `$` to access a column in a data frame.

Notice this returns an unnamed vector.

```{r}
expt$score
```

We can also use `[]` to access the 1st, 2nd, 3rd, etc. column from a data frame.

Notice that this returns a *data frame*.

```{r}
expt[1]

```

## Check the class of the output

```{r}
class(expt[1])
```

## Access a row in the data frame

```{r}
### View the first row
expt[1, ]
```

```{r}
### View the third to the seventh rows
expt[3:7, ]
```

```{r}
### View the third to the seventh rows, but only the third and fourth columns
expt[3:7, 3:4]
```

```{r}
### View the third to the seventh rows, but only the third and fourth columns
expt[3:7, c('score', 'age')]
```

In future courses, we will use special packages to pull out data from data frames. In this course, we'll stick with base-R functionality.

------------------------------------------------------------------------

# VIDEO: Investigating a Data Frame

## How big is the data frame?

```{r}
dim(expt)

# nrow(expt)

# ncol(expt)
```

```{r}
nrow(expt)
```

## Identify the names of the data frame

```{r}
# Option 1
names(expt)

# Option 2
colnames(expt)
```

## View the structure of a data frame

```{r}
str(expt)

### For CodeGrade Assignments
ls.str(expt)
```

-   Note the dimensions

-   Note the `$` by each name.

-   Note the data types (R's best guess)

## Find summary statistics

```{r}
summary(expt)
```

------------------------------------------------------------------------

# VIDEO: Accessing a CSV File

-   Data frames are often stored in CSV files (comma separated values file)

-   How do we load a CSV into R so we can analyze the data frame?

Let's practice:

1.  Download the 'donors.csv' from Brightspace.

2.  Store the file in the folder as this "Getting Started with R.Rmd" Notebook (in your working directory)

## One Way: Navigate "files" pane

Navigate to the folder where the file is.

Select "Import Dataset..."

```{r}
donors
```

## Just be sure to save the dataframe with the name that the assignment says to use.

Suppose an assignment says to call this dataset `donors_2023`:

```{r}
donors_2023 <- donors
```

## A Different Way: Use read.csv()

-   This is the way you'll use on assignments

### What is my working directory?

```{r}
getwd()
```

### Set my working directory

```{r}
setwd("/Users/choughton/Desktop/Eastern/DS550/")
```

## Is the donors.csv file present?

```{r}
list.files()
```

If you don't see the file you need for the assignment, you have to set your working directory (or save the file in the current working directory)

-   Recall an earlier video about setting your default working directory (Go to Tools, then Global Options)

## read.csv()

This approach is what you will use for some assignments in this course.

```{r}
donors_2023 <- read.csv("donors.csv")

# Only works if the CSV file is in your working directory
# Spelling matters
# Quotes matter
# The .csv matters
# Don't open the file in Excel or Numbers
```

Figure out how to do this sooner than rather later!
